\hfb At first, I would like to thank my parents for their support and encouragement. Thanks to them for raising me to be what I am today, believing in me, giving me freedom and space to grow as I deemed fit. A special thanks go to my mother. Without her love and affection, it would have been impossible to finish my work. I also thank my brother and sisters, who have shown their belief in me which gave me constant confidence. Their constant support has often helped me sustain myself throughout. I also take this opportunity to thank one of my teachers Satyendra Singh who has always been helpful in many ways.

\vspace{0.3cm}
\hf At this point, I must thank the faculties at the National Institute of Technology, Rourkela, who are responsible for the academic path that has led me here. I wish to thank Dr. Debajyoti Choudhuri for his unforgettable guidance during my stay at NIT Rourkela. Thanks to Dr. Bikash Sahoo for introducing me to Prof. Swadhin Pattanayak, who deserves a special mention for sharing his philosophy of mathematics and encouragement for the research. 

\vspace{0.3cm}
\hf I must also thank the IISER Kolkata for providing a friendly environment for my academic study. I also acknowledge the staff members and security persons of IISER Kolkata for helping me in numerous ways. The library resources have been tremendously helpful during my whole stay here. I must thank the library staff for this. 

\vspace{0.3cm}
\hf  I am really grateful to the Mathematics Training and Talent Search (MTTS) program as they set the right base, which was helpful for me to build up on. A special thanks to Prof. Kumaresan and Prof. Bhabha Kumar Sharma for their wonderful teaching and ideas, which helped me to handle a research problem. The methodology given by Prof. Kumaresan was always giving me the strength to understand research articles. 

\vspace{0.3cm}
\hf This whole PhD would not have been possible without financial support. I must thank and acknowledge the CSIR-UGC Government of India for this support. 

\vspace{0.3cm}
\hf It is my privilege to express heartful thanks and sincere gratitude to the faculty members of the Department of Mathematics, IISER Kolkata. A special thanks to my RPC members Dr. Sushil Gorai and Dr. Subrata Shyam Roy, for listening to me each year and giving valuable comments. I want to extend my sincere thanks to Satyaki Sir and Shirshendu Sir for their guidance and for being a friend, which made me feel more comfortable. I am also grateful to the non-teaching staff of my department, Adrish Da and Rajesh Da, for their administrative and many other bits of help.

\vspace{0.3cm}
\hf I would like to thank the thesis reviewers, Dr.  Ritwik Mukherjee from NISER Bhubaneswar and Prof. Dr. Janko Latschev from University of Hamburg for their valuable suggestions. They have pointed out corrections and these have been invaluable in improving the mathematical exposition of the thesis.

\vspace{0.3cm} 
\hf I cannot begin to express my thanks to my friends, who played an important role during my stay. Prahllad Da's discussions were beneficial in the early days of my PhD. His experience and knowledge helped me a lot to understand basic tools in algebraic and differential topology. Many thanks to Sandip and Ramyak for the fruitful discussions. When I got stuck on some proofs, conversations with them often solved many problems. I gladly thank Golam, Subrata, Ashish Da and Mrinmoy Da for their availability for any help. Many thanks to Manish, Sanjoy, Jiten, Samiron, Avishek, Sugata. I thank Mukil, Saikat, Sandip, and Anant for their helpful discussions and make valuable comments after reading the thesis. I thank Saikat Panja from IISER Pune for many suggestions and daily discussions. Finally, I would also like to extend my deepest gratitude to Gaurav, Raksha and Sonu for their friendship. This journey would have been tough without you guys.

\vspace{0.3cm}
\hf I thank Anada Marg and my father for teaching me meditation which always gave me positive energy. Thanks to Gurudev for always helping me, showing me the right path and for constant blessings.

\vspace{0.3cm} 
\hf I would also like to thank the creators of beautiful and open source software like \LaTeX, \textsc{Inkscape}, \textsc{GeoGebra} and many more. Also, I should not forget to thank the webpages like Wikipedia, math stack exchange, math overflow and TeX stack exchange. 

\vspace{0.3cm} 
\hf Finally, it comes to the director of the thesis, my supervisor Dr. Somnath Basu. He is the most important person without whom this was absolutely not possible. His way of doing mathematics is totally different. He has crystal clear concepts, geometric intuition and beautiful imagination. He has been very patient and attentive, simultaneously providing guidance and sharing his mathematical insights. I am thankful to him for his guidance and friendly encouragement throughout my work over the last five years. I feel lucky to have him as an advisor who has taken a keen interest in my progress. Apart from mathematics, his contagious enthusiasm and work ethic makes him a great teacher to work under. I could not imagine someone better to learn mathematics from! I also appreciate his patience and efficiency with regular and extensive online discussions during the lockdown period due to the COVID pandemic. I feel fortunate to know him; he will always be an inspiration to me. Thank you so much Sir, for providing a proper guidance, suggestions and feedback throughout my PhD tenure. It was impossible for me to complete this thesis without your support and supervision.

\vspace{0.3cm}
\hf Finally, thanks to all who are not mentioned here but are associated with me.


